---
title: "Regression - Inference"
author: "Howard J"
date: "December 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(UsingR)
library(tidyverse)
```


## Regression Inference

### Inference
$$\frac{\hat \theta - \theta}{\hat \sigma_{\hat \theta}}$$
where,
$\hat \theta $: an estimate of interest  
$\theta$: the estimand of interest  
$\hat \sigma_{\hat \ theta}$: the standard error of $\hat \theta$  


We see that in many cases
such statistics often have the following properties:

Is normally distributed and has a finite sample Student's T distribution if the  variance is replaced with a sample estimate (under normality assumptions).
    3. Can be used to test $H_0 : \theta = \theta_0$ versus $H_a : \theta >, <, \neq \theta_0$.
    4. Can be used to create a confidence interval for $\theta$ via $\hat \theta \pm Q_{1-\alpha/2} \hat \sigma_{\hat \theta}$
    where $Q_{1-\alpha/2}$ is the relevant quantile from either a normal or T distribution.


### Regrssion Model
*Definition:*  
$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$  
where, $\epsilon_i$ is iid Gaussian errors:  
$$\epsilon_i \sim N(0, \sigma^2)$$
$$\hat \beta_0 = \bar Y - \hat \beta_1 \bar X$$
$$\hat \beta_1 = Cor(Y, X) \frac{Sd(Y)}{Sd(X)}$$

*Assumption:* the true model is known.


### Regression Parameters
**Standard Errors**  
$$\sigma_{\hat \beta_1}^2 = Var(\hat \beta_1) = \sigma^2 / \sum_{i=1}^n (X_i - \bar X)^2$$
$$\sigma_{\hat \beta_0}^2 = Var(\hat \beta_0)  = \left(\frac{1}{n} + \frac{\bar X^2}{\sum_{i=1}^n (X_i - \bar X)^2 }\right)\sigma^2$$

