---
title: "Regression - Multivariable"
author: "Howard J"
date: "January 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Basic Regression
#### `lm()` description
`lm()`: fitting linear models, returns an object of class "lm" or for multiple responses of class c("mlm", "lm").  

The functions `summary()` and `anova()` are used to obtain and print a summary and analysis of variance table of the results. The generic accessor functions coefficients, effects, fitted.values and residuals extract various useful features of the value returned by lm.

`r lm()$coefficients`: only provides intercept and coefficient. 
`r summary(lm())$coefficients`: intercept and coefficient, Estimate, Std. Error, t value, Pr(>|t|)

#### predict example
Without 'newdata' data frame, the fitted values are used.
```{r}
x <- rnorm(15)
y <- x + rnorm(15)
predict(lm(y ~ x))
```

```{r}
new <- data.frame(x = seq(-3, 3, 0.5))
predict(lm(y ~ x), new, se.fit = TRUE)
pred.w.plim <- predict(lm(y ~ x), new, interval = "prediction")
pred.w.clim <- predict(lm(y ~ x), new, interval = "confidence")
matplot(new$x, cbind(pred.w.clim, pred.w.plim[,-1]),
        lty = c(1,2,2,3,3), type = "l", ylab = "predicted y")
```




#### Calculation of Coefficients $\beta_i$
```{r}
x <- c(0.10,0.11,0.12,0.13,0.14,0.15,0.16,0.17,0.18,0.20,0.21,0.23)
y <- c(42.0,43.5,45.0,45.5,45.0,47.5,49.0,53.0,50.0,55.0,55.0,60.0)
lm.sol <- lm(y ~ 1 + x)
summary(lm.sol)
```


### Toothpaste
```{r}
### Calculate the confidence interval of the linear model
beta.int <- function(fm,alpha = 0.05){
   A <- summary(fm)$coefficients
   df <- fm$df.residual
   left <- A[,1]-A[,2]*qt(1-alpha/2, df)
   right <- A[,1]+A[,2]*qt(1-alpha/2, df)
   rowname <- dimnames(A)[[1]]
   colname <- c("Estimate", "Left", "Right")
   matrix(c(A[,1], left, right), ncol = 3,
       dimnames = list(rowname, colname ))  
}
```


```{r}
toothpaste <- data.frame(
   X1=c(-0.05, 0.25,0.60,0,   0.25,0.20, 0.15,0.05,-0.15, 0.15,
         0.20, 0.10,0.40,0.45,0.35,0.30, 0.50,0.50, 0.40,-0.05,
        -0.05,-0.10,0.20,0.10,0.50,0.60,-0.05,0,    0.05, 0.55),
   X2=c( 5.50,6.75,7.25,5.50,7.00,6.50,6.75,5.25,5.25,6.00,
         6.50,6.25,7.00,6.90,6.80,6.80,7.10,7.00,6.80,6.50,
         6.25,6.00,6.50,7.00,6.80,6.80,6.50,5.75,5.80,6.80),
   Y =c( 7.38,8.51,9.52,7.50,9.33,8.28,8.75,7.87,7.10,8.00,
         7.89,8.15,9.10,8.86,8.90,8.87,9.26,9.00,8.75,7.95,
         7.65,7.27,8.00,8.50,8.75,9.21,8.27,7.67,7.93,9.26)
)
df <- toothpaste
```

```{r}
lm.sol <- lm(df$Y ~ df$X1 + df$X2)
summary(lm.sol)
```

```{r}
plot(df$Y ~ df$X1)
abline(lm(df$Y ~ df$X1))
```


```{r}
Y <- df$Y
X2 <- df$X2
lm2.sol <- lm(Y ~ X2 + I(X2^2))

x <- seq(min(X2), max(X2), len=200)
y <- predict(lm2.sol, data.frame(X2=x))
plot(Y ~ X2)
lines(x,y)
```

```{r}
lm.new <- update(lm.sol, . ~ .+I(df$X2^2))
summary(lm.new)
beta.int(lm.new)
```


```{r}
lm2.new <- update(lm.new, . ~ .-df$X2)
summary(lm2.new)
```

```{r}
lm3.new <- update(lm.new, . ~ .+df$X1*df$X2)
summary(lm3.new)
```


### Cement
```{r}
cement <- data.frame(
   X1 = c( 7,  1, 11, 11,  7, 11,  3,  1,  2, 21,  1, 11, 10),
   X2 = c(26, 29, 56, 31, 52, 55, 71, 31, 54, 47, 40, 66, 68),
   X3 = c( 6, 15,  8,  8,  6,  9, 17, 22, 18,  4, 23,  9,  8),
   X4 = c(60, 52, 20, 47, 33, 22,  6, 44, 22, 26, 34, 12, 12),
   Y  = c(78.5, 74.3, 104.3,  87.6,  95.9, 109.2, 102.7, 72.5, 
        93.1,115.9,  83.8, 113.3, 109.4)
)

```

```{r}
lm.sol <- lm(Y ~ X1+X2+X3+X4, data = cement)
summary(lm.sol)
```

```{r}
lm.step <- step(lm.sol)
```

```{r}
lm.step$anova
```

```{r}
summary(lm.step)
```


```{r}
drop1(lm.step)
```




### Troubleshooting
#### lm()
Getting Warning: “ 'newdata' had X row but variables found have Y rows” on predict.  

This is a problem of using different names between your data and your newdata and not a problem between using vectors or dataframes.

When you fit a model with the `lm` function and then use `predict` to make predictions, predict tries to find the same names on your newdata. In your first case name x conflicts with mtcars$wt and hence you get the warning.

```r
a <- mtcars$mpg
x <- mtcars$wt
#here you use x as a name
fitCar <- lm(a ~ x) 
#here you use x again as a name in newdata.
predict(fitCar, data.frame(x = mean(x)), interval = "confidence") 
```
       fit      lwr      upr
1 20.09062 18.99098 21.19027

See that in this case you fit your model using the name x and also predict using the name x in your newdata. This way you get no warnings and it is what you expect.


Let's see what happens when I change the name to something else when I fit the model:  
```r
a <- mtcars$mpg
b <- mtcars$wt  # <- name it b this time 
fitCar <- lm(a ~ b) 
predict(fitCar, data.frame(x = mean(x)), interval = "confidence")  # here I am using name x as previously
```
         fit       lwr      upr
1  23.282611 21.988668 24.57655
2  21.919770 20.752751 23.08679
3  24.885952 23.383008 26.38890
4  20.102650 19.003004 21.20230
5  18.900144 17.771469 20.02882
Warning message:
'newdata' had 1 row but variables found have 32 rows 
