

## Discrete Probability Distribution  
### Binomial Distribution  
**Bernoulli and Binomial Random Variables**
Suppose that a trial, or an experiment, whose outcome can be classified as either a success or a failure is performed. If we let $X = 1$ when the outcome is a $success$ and $X = 0$ when it is a $failure$, then the PMF of $X$ is given by:  
$$p(0) = P\{X = 0\} = 1 - p$$  
$$p(1) = P\{X = 1\} = p$$  

where $p$, $0 \le p \le 1$, is the probability that the trial is a *success*.  

A random variable X is said to be a **Bernoulli random variable** if its PMF is given by the equations above for some $p \in (0, 1)$.

Suppose now that $n$ independent trials, each of which results *in a success with probability $p$* and *in a failure with probability $1 − p$*, are to be performed. *If $X$ represents the number of successes that occur in the $n$ trials, then $X$ is said to be a **Binomial random variable** with parameters $(n, p)$*. Thus, *a Bernoulli random variable is just a binomial random variable with parameters (1, p)*.  

The PMF of a binomial random variable having parameters (n, p) is given by:  
$$p(x) = {n \choose x} p^x (1-p)^{(n-x)}$$  
where $x=0,1,...,n$, notice that $x$ begins at 0. ${n \choose 0}$ is 1.


The **Binomial Random Variables** are obtained as the sum of iid **Bernoulli trials**. So if a Bernoulli trial is the result of a coin flip, a binomial random variable is the total number of heads.

To write it out as mathematics, let $X_1,\ldots, X_n$ be iid Bernoulli $(p)$, then $X = \sum_{i=1}^n X_i$ is a binomial random variable. We write out that $X \sim Binomial(n,p)$. The binomial PMF is

$$
P(X = x) =
\left(
\begin{array}{c}
  n \\ x
\end{array}
\right)
p^x(1 - p)^{n-x}
$$

where $x=0,\ldots, n$. Recall that the notation

$$
\left(
  \begin{array}{c} n \\ x \end{array}
\right) = \frac{n!}{x!(n-x)!}
$$

*Note:* $n \choose x$ counts the number of ways of selecting $x$ items out of $n$ *without replacement disregarding the order* of the items. 


*Example:* Suppose a friend has 8 children, $7$ of which are girls and none are twins. If each gender has an independent $50%$ probability for each birth, what's the probability of getting $7$ or more girls out of $8$ births?

```{r}
choose(8, 7) * 0.5^8 + choose(8, 8) * 0.5^8
```
```{r}
pbinom(6, size = 8, prob = 0.5, lower.tail = FALSE)
```
*Notice:* for `lower.tail` argument, if TRUE, probabilities are $P[X \le x]$, if FALSE, $P[X \gt x]$. 




### The Normal Distribution
In fact, the normal distribution only requires two numbers to characterize it. Specifically, a random variable is said to follow a **Normal Distribution** or **Gaussian Distribution** with *mea*n $\mu$ and *variance* $\sigma^2$ if the associated density is:  
$$ (2\pi \sigma^2)^{-1/2}e^{-(x - \mu)^2/2\sigma^2}. $$  

If $X$ is a random variable with this density, then $E[X] = \mu$ and $Var(X) = \sigma^2$, $X$ is denoted as a **Normal Random Variable**. We write as $X\sim N(\mu, \sigma^2)$. When $\mu = 0$ and $\sigma = 1$ the resulting distribution is called **Standard Normal Distribution**. **Standard normal Random Variable** are often labeled $Z$.

If we know that the *population is normally distributed*, we can estimate about the population, by estimating the population mean and variance. (These two population quantities are estimated by the sample mean and the sample variance.)


The most relevant probabilities are.
1. Approximately 68\%, 95\% and 99\%  of the normal density lies within 1, 2 and 3 standard deviations from the mean, respectively.
2. -1.28, -1.645, -1.96 and -2.33 are the $10^{th}$, $5^{th}$, $$2.5^{th}$ and $1^{st}$ percentiles of the standard normal distribution, respectively??
3. By symmetry, 1.28, 1.645, 1.96 and 2.33 are the $90^{th}$, $$95^{th}$, $97.5^{th}$ and $99^{th}$ percentiles of the standard normal distribution, respectively??



### Shifting and Scaling Normal Distribution    
Since the normal distribution is characterized by only the *mean* and *variance*, which are a *shift* and a *scale*, we can transform normal random variables to be standard normals and vice versa. The normal distribution $X \sim N(\mu,\sigma^2)$ can be converted to standard normal distribution:  
$$Z = \frac{X -\mu}{\sigma} \sim N(0, 1^2)$$  

Therefore, normal distribution $X$ can be written as a standard normal times the scale and then plus the shift:    
$$X = \mu + \sigma Z \sim N(\mu, \sigma^2)$$  

Where $X$ is $X \sim N(\mu,\sigma^2)$.  

We can use these facts to answer questions about non-standard normals by relating them back to the standard normal.  


