---
title: "Statistical Inference Notes"
author: "Howard J, based on the work of Caffo"
date: "December 10, 2017"
output:
html_document:
  highlight: pygments
  theme: spacelab
  toc: yes
header-includes: \usepackage{graphicx}
---



## P-values      
### Intro
*Idea:*  
Suppose Null, then how unusual is it to see the estimate we got?  

*Approach:*  
1. Define the *hypothetical distribution of a data summary* (statistic) when "nothing is going on" (*null hypothesis*)  
2. Calculate the summary/statistic with the data we have (*test statistic*)  
3. Compare what we calculated to our hypothetical distribution and see if the value is "extreme" (*p-value*)  

The __P-value__ is the *probability under the null hypothesis of obtaining evidence* as extreme or more extreme than that obtained.


If the P-value is small, it means:  

* $H_0$ is true and we have observed a rare event.  
* Or, $H_0$ is false.  


*Question:* suppose that you get a $t$ statistic of $2.5$ for $15\:df$ testing. What's the probability of getting a $t$ statistic as large as $2.5$?  
$H_0:\mu = \mu_0$  
$H_a : \mu > \mu_0$    
```{r}
pt(2.5, 15, lower.tail = FALSE)
```
Therefore, the probability of seeing evidence as extreme or more extreme than that actually obtained under $H_0$ is `r pt(2.5, 15, lower.tail = FALSE)`


### The Attained Significance Level
In a previous chapter, using a normal test with $n=100$ our test statistic was $2$ for:  
$H_0 : \mu_0  = 30$  
$H_a:\mu > 30$   
Notice that we rejected the one sided test when $\alpha = 0.05$, would we reject if $\alpha = 0.01$, how about $0.001$?  

* *The smallest value for $\alpha$ that you still reject the null hypothesis is called the __attained significance level__*  
* This is equivalent, but philosophically a little different from, the *P-value*.  
* By reporting a P-value the reader can perform the hypothesis test at whatever $\alpha$ level he or she choses.  
* If the P-value is less than $\alpha$ you reject the null hypothesis  
* For two sided hypothesis test, double the smaller of the two one sided hypothesis test P-values  

*Note:* whereas the *P-value* is interpreted in the terms of *how probabilistically extreme* our test statistic is under the null, the *attained significance level* merely conveys what the *smallest level of $\alpha$ that one could reject at*.  

*Revisiting an earlier example:* suppose a friend has $8$ children, $7$ of which are girls and none are twins. If each gender has an independent $50$% probability for each birth, what's the probability of getting $7$ or more girls out of $8$ births?
```{r}
choose(8, 7) * .5 ^ 8 + choose(8, 8) * .5 ^ 8
pbinom(6, size = 8, prob = .5, lower.tail = FALSE)
```

*Poisson Example:* suppose that a hospital has an infection rate of 10 infections per 100 person per day at risk (rate of 0.1) during the last monitoring period. Assume that an infection rate of 0.05 is an important benchmark. Given the model, could the observed rate being larger than 0.05 be attributed to chance?  

* Under $H_0: \lambda = 0.05$
* Consider $H_a: \lambda > 0.05$.  

$\lambda$ - the rate of infections per person/day, $\lambda = 0.05$ means 5 infections per person per 100 days. Thus we want to know whether 9 events per person per 100 days is unusual with respect to a Poisson distribution with a rate of 5 events per 100.  

```{r}
ppois(9, 5, lower.tail = FALSE)
```
