---
title: "Statistical Inference Notes"
author: "Howard J"
date: "December 10, 2017"
output:
html_document:
  highlight: pygments
  theme: spacelab
  toc: yes
header-includes: \usepackage{graphicx}
---

```{r message=FALSE, warning=FALSE}
library(UsingR)
library(tidyverse)
library(grid)
library(png)
```




### Distributions  
For each distribution, R has four functions:  

Prefix  |  Continous  |  Discrete
--------|-------------|-----------
d  | PDF | PMF
p  | CDF | CDF
q  | quantile | quantile
r  | random | random

Distribution root name:  

Distribution | Root
-------------|-----
Binomial     | binom
Poisson      | pois
Normal       | norm  
t            | t  
F            | F  
Chi-square   | chisq  



## Discrete Probability Distribution  
### Binomial Distribution  
#### Bernoulli and Binomial Random Variables  
Suppose that a trial, or an experiment, whose outcome can be classified as either a success or a failure is performed. If we let $X = 1$ when the outcome is a $success$ and $X = 0$ when it is a $failure$, then the PMF of $X$ is given by:  
$$p(0) = P\{X = 0\} = 1 - p$$  
$$p(1) = P\{X = 1\} = p$$  

where $p$, $0 \le p \le 1$, is the probability that the trial is a *success*.  

A random variable X is said to be a **Bernoulli random variable** if its PMF is given by the equations above for some $p \in (0, 1)$.

Suppose now that $n$ independent trials, each of which results *in a success with probability $p$* and *in a failure with probability $1 - p$*, are to be performed. *If $X$ represents the number of successes that occur in the $n$ trials, then $X$ is said to be a **Binomial random variable** with parameters $(n, p)$*. Thus, *a Bernoulli random variable is just a binomial random variable with parameters (1, p)*.  

The PMF of a binomial random variable having parameters (n, p) is given by:  
$$p(x) = {n \choose x} p^x (1-p)^{(n-x)}$$  
where $x=0,1,...,n$, notice that $x$ begins at 0. ${n \choose 0}$ is 1.


The **Binomial Random Variables** are obtained as the sum of iid **Bernoulli trials**. So if a Bernoulli trial is the result of a coin flip, a binomial random variable is the total number of heads.  

To write it out as mathematics, let $X_1,\ldots, X_n$ be iid Bernoulli $(p)$, then $X = \sum_{i=1}^n X_i$ is a binomial random variable. We write out that $X \sim Binomial(n,p)$. The binomial PMF is

$$
P(X = x) =
\left(
\begin{array}{c}
  n \\ x
\end{array}
\right)
p^x(1 - p)^{n-x}
$$

where $x=0,\ldots, n$. Recall that the notation

$$
\left(
  \begin{array}{c} n \\ x \end{array}
\right) = \frac{n!}{x!(n-x)!}
$$

*Note:* $n \choose x$ counts the number of ways of selecting $x$ items out of $n$ *without replacement disregarding the order* of the items.


*Example:* Suppose a friend has 8 children, $7$ of which are girls and none are twins. If each gender has an independent $50%$ probability for each birth, what's the probability of getting $7$ or more girls out of $8$ births?

```{r}
choose(8, 7) * 0.5^8 + choose(8, 8) * 0.5^8
```
```{r}
pbinom(6, size = 8, prob = 0.5, lower.tail = FALSE)
```
*Notice:* for `lower.tail` argument, if TRUE, probabilities are $P[X \le x]$, corresponding to CDF; if FALSE, $P[X \gt x]$, corresponding to survival function.


#### Binomial Distribution   
The __Binomial Distribution__ model deals with finding the *probability of success of an event which has only two possible outcomes in a series of experiments*. For example, tossing of a coin always gives a head or a tail. The probability of finding exactly 3 heads in tossing a coin repeatedly for 10 times is estimated during the binomial distribution.  

Four functions are provided by R:  
`dbinom(x, size, prob)`  
`pbinom(q, size, prob)`  
`qbinom(p, size, prob)`  
`rbinom(n, size, prob)`  

`x, q`: vector of quantiles (input vector of integers), e.g. number of heads.  
`p` is a vector of probabilities.  
`n` is number of observations.  
`size` is the number of trials.  
`prob` is the probability of success of each trial, e.g. the probability for head of each trial.  

**dbinom**  
`dbinom` returns binomial probabilities. The returned value is PMF values at the input vector of integers.  
```{r dbinom}
# Create a sample of 50 numbers which are incremented by 1.
x <- seq(0,50,by = 1)
# Create the binomial distribution.
y <- dbinom(x,50,0.5)
plot(x,y)
```

**pbinom**  
`pbinom(x, size, prob)` gives the cumulative probability distribution of an event. It is a single value representing the cumulative probability.  
*Example:* Probability of getting 26 or less heads from a 51 tosses of a coin.    
```{r pbinom}
x <- pbinom(26,51,0.5)
print(x)
```

*Example:* the cumulative probability distribution for the previous example.  
```{r pbinom2}
x <- seq(0,50,by = 1)
y <- pbinom(x,50,0.5)
plot(x,y)
```

**qbinom**  
`qbinom` takes the cumulative probability value and gives a number which matches this cumulative probability value.  
*Example:* how many heads or less will have a probability of success of 0.25 when a coin is tossed 51 times:  
```{r qbinom}
x <- qbinom(0.25,51,1/2)
print(x)
```

**rbinom**  
`rbinom` generates required number of random values of given probability from a given sample.  
*Example:* generate 8 random values (each value is the number of success, e.g. heads) from a sample of 150 with probability of 0.4:  
```{r rbinom}
x <- rbinom(8,150,.4)
print(x)
```


### The Normal Distribution
In fact, the normal distribution only requires two numbers to characterize it. Specifically, a random variable is said to follow a **Normal Distribution** or **Gaussian Distribution** with *mea*n $\mu$ and *variance* $\sigma^2$ if the associated density is:  
$$ f(x)=(2\pi \sigma^2)^{-1/2}e^{-(x - \mu)^2/2\sigma^2}. $$  
where $-\infty \lt x \lt \infty$  
If $X$ is a random variable with this density, then $E[X] = \mu$ and $Var(X) = \sigma^2$, $X$ is denoted as a **Normal Random Variable**, or simply that $X$ is **Normally Distributed** with parameters $\mu$ and $\sigma^2$. We write as $X\sim N(\mu, \sigma^2)$. When $\mu = 0$ and $\sigma = 1$ the resulting distribution is called **Standard Normal Distribution**. **Standard normal Random Variable** are often labeled $Z$.

If we know that the *population is normally distributed*, we can estimate about the population, by estimating the population mean and variance. (These two population quantities are estimated by the sample mean and the sample variance.)


The most relevant probabilities are:    
1. Approximately 68\%, 95\% and 99\%  of the normal density lies within 1, 2 and 3 standard deviations from the mean, respectively.  
2. -1.28, -1.645, -1.96 and -2.33 are the $10^{th}$, $5^{th}$, $2.5^{th}$ and $1^{st}$ percentiles of the standard normal distribution, respectively??  
3. By symmetry, 1.28, 1.645, 1.96 and 2.33 are the $90^{th}$, $95^{th}$, $97.5^{th}$ and $99^{th}$ percentiles of the standard normal distribution, respectively??  


#### Shifting and Scaling Normal Distribution    
Since the normal distribution is characterized by only the *mean* and *variance*, which are a *shift* and a *scale*, we can transform normal random variables to be standard normals and vice versa. The normal distribution $X \sim N(\mu,\sigma^2)$ can be converted to standard normal distribution:  
$$Z = \frac{X -\mu}{\sigma} \sim N(0, 1^2)$$  

Therefore, normal distribution $X$ can be written as a standard normal times the scale and then plus the shift:    
$$X = \mu + \sigma Z \sim N(\mu, \sigma^2)$$  

Where $X$ is $X \sim N(\mu,\sigma^2)$.  

We can use these facts to answer questions about non-standard normals by relating them back to the standard normal.  


### Exponential Random Variables  
A continuous random variable whose probability density function is given, for some
$λ > 0$, by

$$
f(z) =
\left
\{
  \begin{array}{rcl}
    \overline{\overline{z^2}+\cos z} & \mbox{for} & |z|<3 \\
    0  & \mbox{for} & 3\leq|z|\leq5 \\
    \sin\overline{z} & \mbox{for} & |z|>5
  \end{array}
\right.
$$
